Dark matter haloes: a multistream view

### 2018-03-07 Draft/Chapter5/Paper2017b_Nobmath.tex
initial commit

\documentclass[fleqn,usenatbib,useAMS]{mnras}
%\documentclass[referee,fleqn,usenatbib,useAMS]{mnras}

@@

% \hl{Counting the number of velocity streams in gravitational collapses supplements our knowledge of spatial clustering. Haloes have diverse multistream features within and around. Nonetheless, they host at least one multistream maxima near (at?) the cores of the collapse. }
% 
% We propose a framework to detect potential dark matter haloes in multistream fields. Closed compact regions of the multistream field are detected using local geometrical conditions and properties of the Lagrangian sub-manifold. All the halo particles are guaranteed to be completely outside void regions of the Universe. Majority of the halo candidates are embedded in the largest structure that percolates the entire volume.

%Mysterious dark matter making up about 85\% of all mass in the Universe which is able to cluster plays
%the dominant role in the formation of all observed structures on scales from a fraction to a few hundred of Mpc.
Mysterious dark matter constitutes about 85\% of all mass in the Universe. Clustering of dark matter plays
the dominant role in the formation of all observed structures on scales from a fraction to a few hundreds of Mega-parsecs. Galaxies play a role of lights illuminating these structures so they can be observed. The observations in the last several decades have unveiled opulent geometry of these structures currently known as the cosmic web.
%The web is  assembled from three generic  building blocks: thin curved walls, filaments most of which are crossings of the walls, 
%and knots where several filaments  converge. 
%The web occupies a small volume, the major parts of the volume in the present universe are devoid of the web structures and are known as voids. 
Haloes are the highest concentrations of dark matter and host luminous galaxies. 
%Usually the stars and the gas make a small fraction in the mass budget compared to dark matter mass associated with the galaxies. 
Currently the most accurate modeling of dark matter haloes is achieved in cosmological N-body simulations. Identifying the haloes from the distribution of particles in N-body simulations is one of the problems attracting both considerable interest and efforts.

We propose a novel framework for detecting potential dark matter haloes using the field unique for dark matter  -- multistream field. The multistream field  emerges at the nonlinear stage of the growth of perturbations because the dark matter is collisionless. Counting the number of velocity streams in gravitational collapses supplements our knowledge of spatial clustering. 
%Haloes have diverse multistream features within and around. Nonetheless, they host at least one multistream maxima at the cores of the collapse. }
%We propose to detect potential dark matter haloes in the multistream fields. 
We assume that the virialized haloes have convex boundaries. Closed convex regions of the multistream field are detected by making use of the Hessian of the multistream field. The halo candidates obtained from the multistream field are compared with the FOF and AHF halos.
%All the halo particles are guaranteed to be completely outside void regions of the Universe. Majority of the halo candidates are embedded in the %largest structure that percolates the entire volume.

@@

\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
%-------------------------------------------------------------------------------------------------------------------------------

@@

\hl{The web-like distribution of matter initially revealed by redshift surveys (with less than 300 galaxies by} \citealt{Gregory1978} \hl{and around 1000 galaxies by} \citealt{DeLapparent1986}) \hl{and numerical modeling (using N-body simulations of around 30000 particles by} \citealt{Shandarin1983} and \citealt{Klypin1983a}) \hl{pioneered morphological investigations of the cosmic web structures (see} \citealt{Bond1996}, also reviews by \citealt{Shandarin1989} and \citealt{Weygaert2008}). \hl{Detailed mapping of the Universe has crossed three million objects today, by catalogues such as the Sloan Digital Sky Survey (SDSS;} \citealt{Albareti2016}). The upcoming Large Synoptic Survey Telescope (LSST; \citealt{lsst2009}) \hl{is expected to probe the nature of dark matter using several billion galaxies. On the other hand, cosmological simulations have improved immensely in several aspects -- numerical techniques, parallelization schemes, inclusion of various physical processes, volume and resolution (some of these developments are summarized in }\citealt{Bertschinger1998} and \citealt{Bagla1997}). Modern state-of-the-art simulations like the Illustris Project \cite{Vogelsberger2014} and Q-Continuum \cite{Heitmann2015} \hl{use more than a billion dark matter particles. Finally, the ever improving data analysis techniques have resulted in new and sophisticated density estimators, geometrical and topological indicators. A plethora of algorithms for identifying and characterizing dark matter structures have emerged in last two decades (a summary on cosmological data analysis is highlighted in} \citealt{Weygaert2009a}). \hl{Considering all these improvements, it is worth noting that the proto-structures detected in the modern simulations are qualitatively similar to the quasi-linear description of clustering by Zeldovich Approximation (ZA;} \citealt{Zeldovich1970}). \hl{Location and properties of these structures, i.e., the voids, walls, filaments and haloes maybe inconsistent across different structure finding algorithms, but that is primarily due to varied definitions. }

%Qualitatively the  qualitatively resemble the proto-structures structures  in basically entailed in the quasi-linear description of the ZA theoretical studies \cite{Zeldovich1970}. Although

%All aspects of these studies have been experiencing great advancements in three decades passed since the discovery and first studies of the geometry and topology of the large-scale structures. The galaxy redshift catalogues have grown by thousands of times [by surveys such as Sloan Digital Sky Survey (SDSS; Tegmark et al. 2003; SDSS Collaboration et al. 2016)and the2MASS Redshift Survey (Huchra et al. 2012)] the sizes of cosmological N-body simulations [modern large-scale simulations like Mil- lennium (Springel et al. 2005) and Q-Continuum (Heitmann et al. 2015)] by more than a million times. The number of vari- ous methods for identifying structures has also grown practically from one method2 to several dozens (Colberg et al. 2008, Knebe et al. 2011, Onions et al. 2012, Knebe et al. 2013 and references therein). Measuring or quantifying the structures always has been a difficult problem and many sophisticated techniques both mathe- matically and computationally have been proposed and investigated (see reviews by van deWeygaert & Bond 2008a,b).

%Cosmic web structures have been characterized using several ge- ometrical and topological indicators such as genus curves (Gott, Melott & Dickinson 1986). In an attempt to characterize the shapes of individual regions in the excursion sets of the density field, Sahni, Sathyaprakash & Shandarin (1998) suggested to use partial Minkowski functionals. They developed the method la- belled SURFGEN and applied it to CIC density field obtained in N-body simulations (Sathyaprakash, Sahni & Shandarin 1998; Sheth et al. 2003; Shandarin, Sheth & Sahni 2004). Aragon-Calvo et al. (2007) have developed the multiscale morphology filter detec- tion technique based on the signs of three eigenvalues of the Hessian computed for a set of replicas of the density field filtered on different scales. Similar multiscale approaches to identifying structures are adopted in NEXUS and its extensions to velocity shear, divergence, and tidal fields (Cautun, van de Weygaert & Jones 2013). More recently, persistence and Morse–Smale complexes in the density fields are analysed by Sousbie (2011), Sousbie, Pichon&Kawahara (2011) and Shivshankar et al (2015) to detect multiscale morphol- ogy of the cosmic web.

Most of structure finders are halo finders only and majority of them are stemmed from three underlying algorithms. One of them is the SO (Spherical Over-density) halo finder that defines halos  as spherical regions whose mass density exceeds the mean density by a specified factor \citep{Press1974}. 
Another is the FOF  (Friends-of-Friends) halo finder describing haloes  as the groups of particles separated less than a specified linking length often chosen as 0.2 times the mean particle separation \citep{Davis1985}. 
\hl{ The FOF can be also used for identifying filaments and walls/pancakes by increasing the linking length} (\citealt{Zeldovich1982}, \citealt{Shandarin1983a}, \citealt{Shandarin2010b} )  %(refs in tex file and in additional\_refs.bib)}. 
%\citep{Zeldovich_etal:82,Shandarin:83,Shandarin_etal:10 } 
Finally the DENMAX (DENsity MAXimum) halo finder assumes that the halos are the peaks of the density  fields and thus selects the particles concentrated in the vicinity of the density maxima \citep{Bertschinger1991}. One of the common features of these  techniques is that all three are based on density, in one form or another. And all of them depend on free parameters that are chosen chiefly on the `merits principle' \citep{Forero-Romero2009a} rather than on physics. Over the years all three kinds of the halo finders have been experiencing various modifications and improvements.  A few examples from a long list of these modifications may include:

@@

(v) Computing hierarchical tree of clusters in the phase-space such as the Hierarchical Structure Finder \cite{Maciejewski2009c}, and the 6-D Hierarchical Over-density Tree \cite{Ascasibar2010}. 
(vi) Hybrid algorithms: frameworks such as the Hierarchical Bound-Tracing algorithm \cite{Han2012a} and SURV \cite{Giocoli2010a} identify haloes at multiple time steps from the simulation to find prospective sub-haloes. In addition, there are HOP methods by \cite{Eisenstein1998}, \cite{Tweed2009} and \cite{Skory2010}. %In addition, there are improvised methods to exclude particles that are not gravitationally bound to the halo candidates.

@@

 \hl{ A detailed comparisons of several halo/sub-halo finders is provided in }
%done in projects like 
\cite{Knebe2011a}, \cite{Knebe2013}, \cite{Onions2012}.\hl{ In a nice summary discussing  these developments as well as describing a few new suggestions they  concluded that there was no general consensus for a precise definition of a halo or a sub-halo. Consequently, there were different estimates of number of haloes, halo mass functions, halo centre locations, boundaries and other parameters. }
%A  are given in comparison project papers quoted above.

%In addition to the refinements and generalizations of these basic methods a few new techniques have been developed for identifying haloes and structures with other geometries like filaments and walls.

@@

\hl{For FOF, the free parameter of linking length is generally taken as $b = 0.2$ times the mean separation of particles at $z=0$. This inter-particle separation corresponds to $\Delta_{vir} \approx 180$ if the halo has an isothermal density profile, $\rho \propto r^{-2}$. Using percolation theory,} \cite{More2011} \hl{argued that this corresponds to a rather wide range of over-densities depending on halo mass and density profiles. They found out that $b = 0.2$ corresponds to local over-density $\delta$ within the enclosed halo to be in the range of 250 to 600. Moreover, the resulting FOF-haloes need not have a compact geometry: often the haloes are irregularly shaped, which is unlikely if the halos are virialized.
Hence modern algorithms re-define the halo boundaries by excluding particles using post-processing techniques. In recent simulations with clear delineation of walls and filaments} \cite{Angulo2013a}, \hl{$b=0.05$ was used for finding FOF-haloes since the traditional value of $b = 0.2$ corresponded to structures that percolate into the web structure. }

\hl{Absence of dynamical traits in the FOF and SO algorithms are arguably more crucial. In phase-space, the halo collapse models show collisionless DM particles in oscillatory motions about a core, at successive foldings of the phase-space sheet. The velocity field within each oscillatory spiral is multi-valued in physical space. Incrementing {\it multistream} shells, separated by {\it caustic} surfaces sequentially trace the structures of the cosmic web - walls, filaments and the haloes. Majority of the mass accretion into the haloes along the filaments: from lower multistreams into higher. Thus the DM haloes are not independent of filaments around them, and the hierarchical layers of multistreams represents this portrait precisely. This picture of structure formation was initially theorized using ZA} \cite{Zeldovich1970} and in context of caustics \cite{Arnold1982} as well as in the Adhesion Approximation (\citealt{Gurbatov1989}, \citealt{Kofman1992}). 
\cite{Shandarin1989} reviewed gravitational evolution of density perturbations in the context.

\hl{It has been demonstrated that the multistream field in Eulerian space can be computed directly from the Lagrangian sub-manifold} (\citealt{Shandarin2012} and \citealt{Abel2012b}). About 90\% of the field is single-streaming voids, and the rest of the volume comprises of multistream walls, filaments and haloes. \cite{Ramachandra2015} found the multistream value of $n_{str} \approx 90$ corresponds to virial density $\Delta_{vir} = 200$. On the other hand, DM particles are identified by \citep{Falck2012} as belonging to haloes if they undergo flip-flop along 3 orthogonal axes. \hl{These analyses have opened up a new avenue in studies of halo formation, both qualitative and quantitative. Re-investigations of halo spins, physical radii of the halo, sub-structure in the light of streaming phenomena have shown that the halo structures and formations are more complicated than previously believed.} \cite{Vogelsberger2011} \hl{investigated the distribution of streams in small haloes at various redshifts. They concluded that tracking caustics and streams is better than density, since density fields are noisy in the dense inner regions of haloes. In another study,} \cite{More2015} \hl{argued that the `splashback radius' - distance from the halo core to the first caustic enumerated from outside - is a better physical indicator of DM halo boundary than the virial radius (also see the discussion on turn-around radius of bound objects by} \citealt{Lee2016a}). \cite{Angulo2013a} \hl{also agree with the view that the locally overdense regions correspond better with the volumes within the first caustic than the virialized DM clumps. Recent toy model of anisotropic halo collapse by} \cite{Neyrinck2016} considers intersecting multistream filaments forming spinning nodes. \cite{Ramachandra2017} showed that the virial surfaces of FOF haloes have varying number of streams, including single-streams. Study by \cite{Shandarin2016} delineated the rich sub-structure of haloes using another derivative parameter from the Lagrangian sub-manifold called the `flip-flop' defined on the Lagrangian space.

@@

%Local geometrical analyses on density fields are popularly done to identify structures in N-body simulations. Three-dimensional skeletons of the filaments were traced by \cite{Sousbie2008b} using Hessian eigenvalues on the density field calculated by Cloud-in-Cell (CIC) algorithm. \cite{Aragon-Calvo2007}, \cite{Aragon-Calvo2010} and \cite{Cautun2014a} discussed morphology filters to find clusters in cosmological simulations. \cite{Aragon-Calvo2010} analyse the Hessian eigenvalues of smoothed density field obtained from the Delaunay tessellation field estimator (\citealt{Schaap2000}, \citealt{Weygaert2009a} and \citealt{Cautun2011}). Similar characterization of cosmic web components is done by \cite{Cautun2014a} in density fields using their NEXUS+ algorithm.

%\rhl{Structure/feature detection - like DisPerSE, ORIGAMI}

%Hessian analysis is obviously not limited to density fields. \cite{Hoffman2012a} analysed the differences in the eigenvalue thresholds for Hessians of gravitational potential and velocity shear tensor. The motivation is that the gravitational potential and velocity potential are not identical in the non-linear regime. From the local geometric description of velocity fields, \cite{Hoffman2012a} argued that the cosmic web is better traced by velocity shear tensor than the gravitational field, especially at smaller scales. \cite{Libeskind2013} calculated fractional anisotropy and eigenvectors in the velocity shear tensor field to analyse halo shapes and alignment. Comparison of local geometric analysis of Hessian on various fields like tidal-tensor, velocity-divergence and velocity shear is done in \cite{Cautun2014a}. The eigenvalues of the peculiar gravitational potential or the tidal field tensor the are investigated by \cite{Hahn2007}, \cite{Forero-Romero2009a}, \cite{Hoffman2012a} and \cite{Cautun2014a}. To classify the structures into void, wall, filament and cluster, \cite{Hahn2007} checked the number of eigenvalues greater than 0 at each point in the Simulation box. Analysis of a non-zero eigenvalue thresholds on the same Hessian is done by \cite{Forero-Romero2009a}, where they argued that volume fraction of the void is higher for positive values of the thresholds. They also concluded that smoothing scales do not play an important role in the volume fractions of the classified structures.

%There is an increasing interest in the local geometrical measures for detecting filaments in large astronomical surveys. \cite{Sousbie2008a} detected skeleton of filaments of the Sloan Digital Sky Survey (SDSS) and compared to the corresponding galaxy distribution. In smoothed density of mock galaxy distribution, \cite{Bond2010a} studied the projection of eigenvalues. The Hessian eigenvector corresponding to the largest eigenvalue is used by \cite{Bond2010b} \hl{to trace individual filaments in N-body simulations and the SDSS redshift survey data.}

In this paper, we identify potential haloes by utilizing multistreaming as the governing dynamical phenomena. Review of the DM particle clustering in a one-dimensional dimensional universe is made in Section \ref{sec:HaloFormation}, with a description of multistream fields. The cosmological simulation is described in Section \ref{sec:simulation}. In Section \ref{sec:haloDetection}, we prescribe a framework to identify dark matter haloes using Hessian eigenvalues in the multistream field. We also discuss a few properties of the haloes. Comparison of the multistream haloes with AHF and FOF algorithms is done in \ref{sub:compareHalo}. We also discuss the spatial distribution of the dark matter haloes along the percolating structure.
%

@@

\hl{New section}

\begin{figure*} 
\centering\includegraphics[height=7cm]{fig1.pdf} 
\caption{Dynamical collapse of dark matter in one-dimensional universe: Top panels show the $(\bmath{p}, \bmath{x})$ phase-space manifold of the dark matter sheet at redshifts $z_1$, $z_2$, $z_3$ and $z = 0$. Dots represent the dark matter particles. The momentum values are chosen at arbitrary scales. Bottom panels show the corresponding multistream field multistream field $n_{str}(\bmath{x}, z)$ (red) and density field $\rho(\bmath{x}, z)$ (gray). }
\label{fig:1d}
\end{figure*}

We begin with a simple illustration showing the formation of a few haloes in a one-dimensional simulation. Dark matter clustering in a (1+1)-dimensional phase-space $(\bmath{p}, \bmath{x})$ (where $p$ is the momentum and $x$ is the co-moving Eulerian coordinate) at four successive time steps is shown in the top panels of \autoref{fig:1d}. The lower panels show the corresponding multistream field (\citealt{Shandarin2012} and \citealt{Abel2012b}) $n_{str}(\bmath{x}, z)$ (red) and density field $\rho(\bmath{x}, z)$ (gray). At $z_1$ (left-most panel), velocity is single-valued in Eulerian co-ordinates shown, except at a small three-stream region near $\bmath{x} = 5\pi/4$. This is the first instance of multistreaming in the region, which was previously had $n_{str} = 1$ throughout. The interface of $n_{str} = 1$ and $n_{str} = 3$ regions is also the location of the first caustic. On the other hand, the density calculated at a high resolution shows variations, even in the mono-streaming regions. The variations are especially more pronounced around  the caustic (near $\bmath{x} = 5\pi/4$).

@@

The right-most panel in \autoref{fig:1d} shows the final structure at $z=0$. Two large spirals have spatially merged. These collapse environments are naturally very complex, with an increased number of successive caustic formation and merging. 
The corresponding velocity streams also show a more complicated structure. Clearly, the multistream field has a saddle point that is not as low as $n_{str} = 1$. %\ghl{The one-dimensional  does not demonstrate this clearly. The highest peak on the left at $\sim 3\pi$/2 may jump from 5 to 9 but the red line too thick to see this. We need to discuss this. - REMOVED that line. Better?}
This poses a bigger problem in the context of most of halo detection algorithms, and we discuss this in Appendix \ref{appendix:Eigen}.

%$3 \supseteq  5 \supseteq  7 \supseteq  9 \supseteq  11 \ldots$.
\subsection{Collapse in higher dimensions}

@@

The collapse of a uniform ellipsoid also results in a simultaneous collapse of the entire ellipsoid
however this time not into a point but into a two-dimensional ellipse \citep{Lin1965}.% etal 1965) %\cite{Lin_etal:65}
Another customarily used spherical model of halo formation by \cite{Fillmore1984} and \cite{Bertschinger1985} does not consider
the initial collapse at all. Instead it assumes self-similar initial conditions and the halo at advanced stage with formally infinite number of spherical caustic shells.

@@

Caustic formations and mass accretion are also seen to occur more along the higher streams, which makes the haloes non-spherical, with the alignment generally determined by a complicated interplay  of the intensities of the streams from neighboring filamentary structures. Number of streams corresponding to the dark matter halo also has a local environment dependence. The three small haloes in \autoref{fig:full}, where the number of streams are higher than the neighbouring filaments, are aligned along three intersecting filaments. Halo environment studied in \cite{Ramachandra2015} show similar hierarchical variation in $n_{str}$ values. %\ghl{[I do not clearly understand what it means] - EDITED, better?}. 
The halo environments are thus very complicated, and empirical thresholds (on multistream or density fields) may not account for all the haloes uniformly. Hence we use a local geometrical method to identify potential haloes in multistream fields.

@@

%The multistream field can be computed at arbitrary resolutions of diagnostic grids. The parameter `refinement factor' denotes the ratio of separation of the particles in Lagrangian grid, $l_l$, to side length of diagnostic grid $l_d$. In a simulation of $128^3$ particles, for instance, multistream field computed on a diagnostic grid of size $256^3$ would have a refinement factor of $l_l/l_d = 2$. The effects of refinement of the diagnostic grid are discussed in Section \ref{sec:voids} and Section \ref{sub:HaloProperties}.

\begin{figure} 
\centering\includegraphics[width=8cm]{fig2.pdf} 
\caption{Multistream field contours: The multistream field is calculated at 8 times the native resolution. Each inner convex blobs (red) surround local multistream maxima inside. Surrounding outer shell(blue) is not convex throughout the surface, and the outermost gray multistream surface displays a filamental geometry.}
\label{fig:full}
\end{figure}

@@

%\ghl{I think that the halo finder algorithm should be explained here including the def. of Hessian. Otherwise the term 'Hessian' - DONE HERE used before it is defined in Sec. 4.1. Explaining the refinement factor in more detail would be also usfull. - DONE with simulations}

\hl{New sub-section}

Hessian matrix $\mathbfss{H}(f)$ of a scalar field $f$ involves local second-order variations in three orthogonal directions. Each element of the Hessian matrix $\mathbfss{H}_{ij}(f)$ (where $i$ and $j$ can be any of $x$, $y$ or $z$ directions) is given by \autoref{eq:Hess}.

\begin{equation}
\label{eq:Hess}
 \mathbfss{H}_{ij}(f) = \frac{\partial^2 f}{\partial x_i \partial x_j}
\end{equation}

@@

% Venn stuff removed from here, and moved towards the end

>>>>

\begin{comment}
\documentclass[fleqn,usenatbib,useAMS]{mnras}
%\documentclass[referee,fleqn,usenatbib,useAMS]{mnras}

@@

Mysterious dark matter constitutes about 85\% of all mass in the Universe. Clustering of dark matter plays the dominant role in the formation of all observed structures on scales from a fraction to a few hundreds of Mega-parsecs. Galaxies play a role of lights illuminating these structures so they can be observed. The observations in the last several decades have unveiled opulent geometry of these structures currently known as the cosmic web. Haloes are the highest concentrations of dark matter and host luminous galaxies. Currently the most accurate modeling of dark matter haloes is achieved in cosmological N-body simulations. Identifying the haloes from the distribution of particles in N-body simulations is one of the problems attracting both considerable interest and efforts. We propose a novel framework for detecting potential dark matter haloes using the field unique for dark matter  -- multistream field. The multistream field emerges at the nonlinear stage of the growth of perturbations because the dark matter is collisionless. Counting the number of velocity streams in gravitational collapses supplements our knowledge of spatial clustering. We assume that the virialized haloes have convex boundaries. Closed and convex regions of the multistream field are hence isolated by imposing a positivity condition on all three eigenvalues of the Hessian estimated on the smoothed multistream field. In a single-scale analysis of high multistream field resolution and low softening length, the halo substructures with local multistream maxima are isolated as individual halo sites.

%The halo candidates obtained from the multistream field are compared with the FOF and AHF halos.

@@

\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\end{comment}
%-------------------------------------------------------------------------------------------------------------------------------
\chapter{Dark matter haloes: a multistream view}

@@

The web-like distribution of matter initially revealed by redshift surveys (with less than 300 galaxies by \citealt{Gregory1978} and around 1000 galaxies by \citealt{DeLapparent1986}) and numerical modeling (using N-body simulations of around 30000 particles by \citealt{Shandarin1983} and \citealt{Klypin1983a}) pioneered morphological investigations of the cosmic web structures (see \citealt{Bond1996}, also reviews by \citealt{Shandarin1989} and \citealt{Weygaert2008}). Detailed mapping of the Universe has crossed three million objects today, by catalogues such as the Sloan Digital Sky Survey (SDSS; \citealt{Albareti2016}). The upcoming Large Synoptic Survey Telescope (LSST; \citealt{lsst2009}) is expected to probe the nature of dark matter using several billion galaxies. On the other hand, cosmological simulations have improved immensely in several aspects -- numerical techniques, parallelization schemes, inclusion of various physical processes, volume and resolution (some of these developments are summarized in \cite{Bertschinger1998} and \cite{Bagla1997}. Modern state-of-the-art simulations like the Illustris Project \cite{Vogelsberger2014}, the EAGLE project \cite{Schaye2015} and Q-Continuum \cite{Heitmann2015} use more than a billion dark matter particles. Finally, the ever improving data analysis techniques have resulted in new and sophisticated density estimators, geometrical and topological indicators. A plethora of algorithms for identifying and characterizing dark matter structures have emerged in last two decades (a summary on cosmological data analysis is highlighted in \citealt{Weygaert2009a}). Considering all these improvements, it is worth noting that the proto-structures detected in the modern simulations are qualitatively similar to the quasi-linear description of clustering by Zeldovich Approximation (ZA; \citealt{Zeldovich1970}). Location and properties of these structures, i.e., the voids, walls, filaments and haloes maybe inconsistent across different structure finding algorithms, but that is primarily due to varied definitions.

Most of structure finders are halo finders only and majority of them are stemmed from three underlying algorithms. One of them is the SO (Spherical Over-density) halo finder that defines halos  as spherical regions whose mass density exceeds the mean density by a specified factor \citep{Press1974}. 
Another is the FOF  (Friends-of-Friends) halo finder describing haloes  as the groups of particles separated less than a specified linking length often chosen as 0.2 times the mean particle separation \citep{Davis1985}. 
The FOF can be also used for identifying filaments and walls/pancakes by increasing the linking length (\citealt{Zeldovich1982}, \citealt{Shandarin1983a}, \citealt{Shandarin2010b}). Finally the DENMAX (DENsity MAXimum) halo finder assumes that the halos are the peaks of the density  fields and thus selects the particles concentrated in the vicinity of the density maxima \citep{Bertschinger1991}. One of the common features of these  techniques is that all three are based on density, in one form or another. And all of them depend on free parameters that are chosen chiefly on the `merits principle' \citep{Forero-Romero2009a} rather than on physics. Over the years all three kinds of the halo finders have been experiencing various modifications and improvements.  A few examples from a long list of these modifications may include:

@@

(v) Computing hierarchical tree of clusters in the phase-space such as the Hierarchical Structure Finder \cite{Maciejewski2009c}, and the 6-D Hierarchical Over-density Tree \cite{Ascasibar2010}. 
(vi) Hybrid algorithms: frameworks such as the Hierarchical Bound-Tracing algorithm \cite{Han2012a} and SURV \cite{Giocoli2010a} identify haloes at multiple time steps from the simulation to find prospective sub-haloes. In addition, there are HOP methods by \cite{Eisenstein1998}, \cite{Tweed2009} and \cite{Skory2010}.

@@

A detailed comparisons of several halo/sub-halo finders is provided in \cite{Knebe2011a}, \cite{Knebe2013}, \cite{Onions2012}. In a nice summary discussing  these developments as well as describing a few new suggestions they  concluded that there was no general consensus for a precise definition of a halo or a sub-halo. Consequently, there were different estimates of number of haloes, halo mass functions, halo centre locations, boundaries and other parameters.

@@

For FOF, the free parameter of linking length is generally taken as $b = 0.2$ times the mean separation of particles at $z=0$. This inter-particle separation corresponds to $\Delta_{vir} \approx 180$ if the halo has an isothermal density profile, $\rho \propto r^{-2}$. Using percolation theory, \cite{More2011} argued that this corresponds to a rather wide range of over-densities depending on halo mass and density profiles. They found out that $b = 0.2$ corresponds to local over-density $\delta$ within the enclosed halo to be in the range of 250 to 600. Moreover, the resulting FOF-haloes need not have a compact geometry: often the haloes are irregularly shaped, which is unlikely if the halos are virialized.
Hence modern algorithms re-define the halo boundaries by excluding particles using post-processing techniques. In recent simulations with clear delineation of walls and filaments \cite{Angulo2013a}, $b=0.05$ was used for finding FOF-haloes since the traditional value of $b = 0.2$ corresponded to structures that percolate into the web structure.

Absence of dynamical traits in the FOF and SO algorithms are arguably more crucial. In phase-space, the halo collapse models show collisionless DM particles in oscillatory motions about a core, at successive foldings of the phase-space sheet. The velocity field within each oscillatory spiral is multi-valued in physical space. Incrementing {\it multistream} shells, separated by {\it caustic} surfaces sequentially trace the structures of the cosmic web - walls, filaments and the haloes. Majority of the mass accretion into the haloes along the filaments: from lower multistreams into higher. Thus the DM haloes are not independent of filaments around them, and the hierarchical layers of multistreams represents this portrait precisely. This picture of structure formation was initially theorized using ZA \cite{Zeldovich1970} and in context of caustics \cite{Arnold1982} as well as in the Adhesion Approximation (\citealt{Gurbatov1989}, \citealt{Kofman1992}). 
\cite{Shandarin1989} reviewed gravitational evolution of density perturbations in the context.

It has been demonstrated that the multistream field in Eulerian space can be computed directly from the Lagrangian sub-manifold (\citealt{Shandarin2012} and \citealt{Abel2012}). About 90\% of the field is single-streaming voids, and the rest of the volume comprises of multistream walls, filaments and haloes. \cite{Ramachandra2015} found the multistream value of $n_{str} \approx 90$ corresponds to virial density $\Delta_{vir} = 200$. On the other hand, DM particles are identified by \citep{Falck2012} as belonging to haloes if they undergo flip-flop along 3 orthogonal axes. These analyses have opened up a new avenue in studies of halo formation, both qualitative and quantitative. Re-investigations of halo spins, physical radii of the halo, sub-structure in the light of streaming phenomena have shown that the halo structures and formations are more complicated than previously believed. \cite{Vogelsberger2011} investigated the distribution of streams in small haloes at various redshifts. They concluded that tracking caustics and streams is better than density, since density fields are noisy in the dense inner regions of haloes. In another study, \cite{More2015} argued that the `splashback radius' - distance from the halo core to the first caustic enumerated from outside - is a better physical indicator of DM halo boundary than the virial radius (also see the discussion on turn-around radius of bound objects by \citealt{Lee2016a}). \cite{Angulo2013a} also agree with the view that the locally overdense regions correspond better with the volumes within the first caustic than the virialized DM clumps. Recent toy model of anisotropic halo collapse by \cite{Neyrinck2016} considers intersecting multistream filaments forming spinning nodes. \cite{Ramachandra2017} showed that the virial surfaces of FOF haloes have varying number of streams, including single-streams. Study by \cite{Shandarin2016} delineated the rich sub-structure of haloes using another derivative parameter from the Lagrangian sub-manifold called the `flip-flop' defined on the Lagrangian space.

@@

In this paper, we identify potential haloes by utilizing multistreaming as the governing dynamical phenomena. A review of the DM particle clustering in a one-dimensional dimensional universe is made in Section \ref{sec:HaloFormation}, and the concept of multistream field is extended to higher dimensions. The multistream field is computed on the cosmological simulations described in Section \ref{sec:simulation}. The halo identification framework in this field is described in Section \ref{sec:haloDetection}. This algorithm isolates convex regions of the multistream field using Hessian eigenvalues, each enclosing a local multistream maximum. Without employing any non-local thresholds that several halo finders generally use, these convex multistream regions are identified as potential halo sites. We also illustrate the significance of multistream refinement and softening scales in finding subhaloes. However, this paper does not focus on adaptive multi-scale analyses for substructure studies. A few properties of the multistream haloes are discussed in Section \ref{sub:HaloProperties}, and comparison of these haloes with AHF and FOF algorithms is done in Section \ref{sub:compareHalo}. We also discuss the spatial distribution of the dark matter haloes along the percolating web structure.

@@

\begin{figure*} 
\centering\includegraphics[height=7cm]{Chapter5/Source_v2/fig1.pdf} 
\caption{Dynamical collapse of dark matter in one-dimensional universe: Top panels show the $(\mathbf{p}, \mathbf{x})$ phase-space manifold of the dark matter sheet at redshifts $z_1$, $z_2$, $z_3$ and $z = 0$. Dots represent the dark matter particles. The momentum values are chosen at arbitrary scales. Bottom panels show the corresponding multistream field multistream field $n_{str}(\mathbf{x}, z)$ (red) and density field $\rho(\mathbf{x}, z)$ (gray). }
\label{fig:1d}
\end{figure*}

We begin with a simple illustration showing the formation of a few haloes in a one-dimensional simulation. Dark matter clustering in a (1+1)-dimensional phase-space $(\mathbf{p}, \mathbf{x})$ (where $p$ is the momentum and $x$ is the co-moving Eulerian coordinate) at four successive time steps is shown in the top panels of \autoref{fig:1d}. The lower panels show the corresponding multistream field (\citealt{Shandarin2012} and \citealt{Abel2012}) $n_{str}(\mathbf{x}, z)$ (red) and density field $\rho(\mathbf{x}, z)$ (gray). At $z_1$ (left-most panel), velocity is single-valued in Eulerian co-ordinates shown, except at a small three-stream region near $\mathbf{x} = 5\pi/4$. This is the first instance of multistreaming in the region, which was previously had $n_{str} = 1$ throughout. The interface of $n_{str} = 1$ and $n_{str} = 3$ regions is also the location of the first caustic. On the other hand, the density calculated at a high resolution shows variations, even in the mono-streaming regions. The variations are especially more pronounced around  the caustic (near $\mathbf{x} = 5\pi/4$).

@@

The right-most panel in \autoref{fig:1d} shows the final structure at $z=0$. Two large spirals have spatially merged. These collapse environments are naturally very complex, with an increased number of successive caustic formation and merging. 
The corresponding velocity streams also show a more complicated structure. Clearly, the multistream field has a saddle point that is not as low as $n_{str} = 1$. This poses a bigger problem in the context of most of halo detection algorithms, and we discuss this in Appendix \ref{appendix:Eigen}.

\subsection{Collapse in higher dimensions}

@@

The collapse of a uniform ellipsoid also results in a simultaneous collapse of the entire ellipsoid
however this time not into a point but into a two-dimensional ellipse (\citealt{Lin1965}, \citealt{Icke1975}, \citealt{Eisenstein1995}).
Another customarily used spherical model of halo formation by \cite{Fillmore1984} and \cite{Bertschinger1985} does not consider
the initial collapse at all. Instead it assumes self-similar initial conditions and the halo at advanced stage with formally infinite number of spherical caustic shells.

@@

Caustic formations and mass accretion are also seen to occur more along the higher streams, which makes the haloes non-spherical, with the alignment generally determined by a complicated interplay  of the intensities of the streams from neighboring filamentary structures. Number of streams corresponding to the dark matter halo also has a local environment dependence. The three small haloes in \autoref{fig:full}, where the number of streams are higher than the neighbouring filaments, are aligned along three intersecting filaments. Halo environment studied in \cite{Ramachandra2015} show similar hierarchical variation in $n_{str}$ values. The halo environments are thus very complicated, and empirical thresholds (on multistream or density fields) may not account for all the haloes uniformly. Hence we use a local geometrical method to identify potential haloes in multistream fields.

@@

\begin{figure} 
\centering\includegraphics[width=8cm]{Chapter5/Source_v2/fig2.pdf} 
\caption{Multistream field contours: The multistream field is calculated at 8 times the native resolution. Each inner convex blobs (red) surround local multistream maxima inside. Surrounding outer shell(blue) is not convex throughout the surface, and the outermost gray multistream surface displays a filamental geometry.}
\label{fig:full}
\end{figure}

@@

Hessian matrix $\mathbf{H}(f)$ of a scalar field $f$ involves local second-order variations in three orthogonal directions. Each element of the Hessian matrix $\mathbf{H}_{ij}(f)$ (where $i$ and $j$ can be any of $x$, $y$ or $z$ directions) is given by \autoref{eq:Hess}.

\begin{equation}
\label{eq:Hess}
 \mathbf{H}_{ij}(f) = \frac{\partial^2 f}{\partial x_i \partial x_j}
\end{equation}

@@

The halo mass functions from all three finders are shown in \autoref{fig:hmf}. For smaller haloes of order of $10^{13} M_{\odot}$, our method predicts a slightly higher number of haloes than FOF and AHF. For the most massive haloes of mass around $10^{14} M_{\odot}$, number of $\lambda_3$-haloes is fewer than the other 2 methods, albeit around the range of error of AHF-haloes.
